Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-05 parallel.tex
Edits to the overview section. Trying to unify notations, so several other sections are also affected.

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained on a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could be
either a core or a node, depending on details of the system.  Note
that the current version of Enzo is not threaded, although a hybrid
MPI + OpenMP version is currently under development.}.  
Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors, $N_p$.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state in general terms which routines within the
code will be most costly during a given simulation.  As such, we have
designed a lightweight registering system that has been implemented
for the most commonly used routines (such as the hydrodynamic and
gravity solvers) as well as refinement level timers that measure the
time spent on each level.  Beyond this minimal set of routines, we
have designed a simple way for the user to modify the source by adding
\texttt{TIMER\_START("Your\_Routine\_Name")} and
\texttt{TIMER\_END("Your\_Routine\_Name")}.  These timers are created
and managed individually on each processor in an asynchronous fashion,
and contribute minimal computational, memory, and IO overhead.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_levels.eps}
\hfill
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_routines.eps}
\end{center}
\caption{\emph{Left:} Strong scaling test of a 512$^3$ AMR
cosmological calculation.  The root grid scaling is not representative
of the true strong scaling of Enzo because the root grid tiles are not
repartitioned when a simulation is restarted with additional cores.
The weak scaling test in Figure \ref{fig.uniscale} is more
representative of the scaling on the root level.  The performance in
the refined levels show good strong scaling up to 2,048 cores.
\emph{Right:} Time spent in representative major routines in the same
AMR calculation.}
\label{fig:strong_scaling}
\end{figure}

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{2011ApJ...738...54K}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling to high core counts.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
Enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in Enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with fewer
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved more quickly, and the
opposite occurs with larger simulations.

>>>>

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained on a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could be
either a core or a node, depending on details of the system.  Note
that the current version of \enzo\ is not threaded, although a hybrid
MPI + OpenMP version is currently under development.}.  
Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors, $N_p$.

@@

Because of the wide variety of simulations, methods, and uses of \enzo,
it is difficult to state in general terms which routines within the
code will be most costly during a given simulation.  As such, we have
designed a lightweight registering system that has been implemented
for the most commonly used routines (such as the hydrodynamic and
gravity solvers) as well as refinement level timers that measure the
time spent on each level.  Beyond this minimal set of routines, we
have designed a simple way for the user to modify the source by adding
\texttt{TIMER\_START("Your\_Routine\_Name")} and
\texttt{TIMER\_END("Your\_Routine\_Name")}.  These timers are created
and managed individually on each processor in an asynchronous fashion,
and contribute minimal computational, memory, and IO overhead.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_levels.eps}
\hfill
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_routines.eps}
\end{center}
\caption{\emph{Left:} Strong scaling test of a 512$^3$ AMR
cosmological calculation.  The root grid scaling is not representative
of the true strong scaling of \enzo\ because the root grid tiles are not
repartitioned when a simulation is restarted with additional cores.
The weak scaling test in Figure \ref{fig.uniscale} is more
representative of the scaling on the root level.  The performance in
the refined levels show good strong scaling up to 2,048 cores.
\emph{Right:} Time spent in representative major routines in the same
AMR calculation.}
\label{fig:strong_scaling}
\end{figure}

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{2011ApJ...738...54K}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using \enzo\ in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling to high core counts.

@@

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
\enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in \enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with fewer
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved more quickly, and the
opposite occurs with larger simulations.
